### 2018-08-02 Writing/disparity_review.tex
added placeholder for Natalie's comment
For exporting the manuscript to google docs
Typo in title
Restructured paper INIT

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: pest practices for disparity analysis.

@@

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------
\section{Introduction}

@@

\begin{itemize}
    \item{Which data should we use in disparity analyses?}
    %[One line abstract of Melanie's paper on inapplicable data].
    Furthermore, some simple methods already exist to account for caveats in data. 
    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
    % NC: This item currently doesn't seem to actually give advice about which data to use...
    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
    [One line abstract of Sylvain's paper on morphospaces plottings].
    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
    One solution to this is to use more simple 2D/3D ordinations.
    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
    \item{Which disparity metrics should we use?}
    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g. in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
    \item{How should we test disparity hypotheses?}
    Finally, it is possible to use non-parametric methods to test specific disparity questions.
    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
\end{itemize}

>>>>

%Running head
\begin{flushright}
Version dated: \today
\end{flushright}
\bigskip
\noindent RH: best practices for disparity analysis.

@@

%---------------------------------------------
%
%       NEW PLAN
%
%---------------------------------------------

% Intro
% IN -> history
%     -Disparity or diversity?
%     Is disparity just diversity? No.
% OU -> Disparity can be many things

% -Diversity of disparities
% IN -> There are many different disparity
%     -Ecological
%     -Functional
%     -Phylogenetic

%     However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?
%     Maybe. Disparity traits need to be carefully selected in order to properly reflect the thing. Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)

% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)

% - Trait disparity
% IN -> If it's just traits, can all of the traits be used?
%     Which traits should be collected?
%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
%     Can disparity analysis work with any kind of trait?
%     Does the part represent the whole?
% OU -> And how do these traits need to be analysed?

% - Disparity is multidimensional
% IN -> Disparity is multidimensional by essence
%     Do we need trait reduction (ordination) or can we do without?
%     What do we measure from these multidimensional spaces?
% OU -> And how do we test these things?

% - Disparity hypothesis
% IN -> So how do we go for testing hypotheses in disparity?
%     What are our null expectations?
%     How can we analyse it through time?

% - A disparity future
% IN -> The definition of disparity and its uses have broadened since the early disparity start.
%     What can we get from other fields to make disparity better
%     But also what can we bring to other fields (discussions with Abigail)

%---------------------------------------------
%
%       INTRODUCTION
%
%---------------------------------------------

\section{Introduction}

%Historical blurb
Variation in the shape, form and function of organisms often follows discrete patterns rather than continuous gradients.
These patterns can be pictured as multidimensional spaces encompassing all the possible forms an organism can take, and where each individual region of this space is the result of a discrete combination of many variables.
One commonly used approach to analyse these discrete patterns, and to understand the processes that lead to them, is to analyse the disparity of the morphological characteristics of the group, often referred to as ``morphological disparity analysis''.

The initial informal definition of disparity was ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (REFS).
Currently, however, there is no consensus on whether we should restrict use of the term disparity to this initial definition, or whether it should be used as a broader umbrella term for studies of multivariate variation. 
Here we use a broader definition and suggest that disparity is a pattern of variation.
Throughout this review, we aim to highlight the diversity of ``disparity'' analyses and how they can aid our understanding of the structure of biodiversity.

\subsection{Is disparity equivalent to diversity?}
One obvious aspect when measuring biodiversity is the measurement of raw species diversity (at any scale; $\alpha$, $\beta$, $\gamma$).
Such biodiversity metric intuitively describes the amount of reproductive isolated populations in a system.
On the other hand, disparity aims at describing the amount of measured traits in a system (see below), in such a way, a high morphological disparity value is a measurement of a high diversity of morphologies.
The link between both disparity and diversity has been scrutinised under different scenarios with most conclusions supporting an effective decoupling between both [CITE the disparity/diversity papers].
This is intuitively understandable when comparing the both metrics for two contrasted groups of organisms, for example, rodents would display a high diversity (@@@ sp) with a relatively low disparity (a majority of rodents have the typical rat shape) whereas carnivores would display a relatively lower diversity (@@@ sp) with a higher disparity (with body shapes ranging from a weasel to an elephant seal).
However, it is important to note that in more nuanced groups, diversity and disparity can be link where a really low diversity would tend to yield a low disparity and vice versa.

%ii) the ensuing ideas on evolvability
2 - Disparity can be linked to evolvability? A morphological disparate group is likely to occupy more niches (but see the debate for the link between traits and ecological niches [CITES]) and more likely to be the result from character displacement and/or competition.
Also, more disparate groups are more likely to lead to disparate radiations as different ecological strategies get explored.

% OU -> Disparity can be many things

\section{A diversity of disparities}
% IN -> There are many different disparity
This diversity (and disparity) of methods to analysis disparity provides biologists with a powerful and comprehensive toolkit for understanding the evolution of biological diversity.
However, as with any great power, it comes with great responsibility. 
With the expansion of methodological approaches in the last decade, we identify three main issues that could reduce the power of disparity analyses if they are not tackled.
First, many recent studies have lost focus; they are no longer hypothesis-driven, instead they are undertaken to characterise biological variation for its own sake.
Second, even where there is a clear question being explored, the data being used for disparity analyses is inappropriate and often driven by the availability of large datasets rather than the question being asked.
Third, the methods being used to analyse these datasets may not always be appropriate for the question and/or data at hand.
Here we deal with some of these issues and present best practice guidelines to allow researchers to optimise their disparity analysis.

\subsubsection{Morphological disparity}

\subsubsection{Functional disparity}
3.2- Functional disparity: seems to work at any scale and can be used to test convergence (species with similar trait combination can have similar functions) or the link between form and function (Diaz paper). However, the link between trait and function can sometimes be hard to grasp since traits can be linked to multiple functions and multiple functions can be linked to a single trait.
\subsubsection{Ecological disparity}
3.1- Ecological disparity: promising tool at a macro-evolutionary scales (but maybe less at micro-evolutionary scales). E.g. the concept of ecological guilds can be distinguished in the fossil record (carnivoran guild etc.) but it might be more difficult to grasp at a micro-scale level (CITE plant trait/niche papers).
\subsubsection{Phylogenetic disparity}

%However, do these disparities represent anything more than the disparity of those traits rather than the disparity of the thing?

%Maybe. Disparity traits need to be carefully selected in order to properly reflect the thing. Disparity is only a proxy to describe the pattern and does not necesserly make conclusions on the process (but see disparity pattern vs disparity process)

% OU -> But in general, disparity is a disparity of traits (Best practice: by the way, it's good to have a clear selection of the traits)

\section{Trait disparity}
% IN -> If it's just traits, can all of the traits be used?
Once one has a clear question, it is important to collect the appropriate data for that question to avoid some major caveats.
For example, if a study focuses on the ecological adaptive radiation of a group, ecological traits are crucial and should be preferred over other kinds of traits, such as those used for spec).ies delimitation.
Tangentially, more data are not necessarily better, especially when the data were originally collected for different questions (e.g. cladistic data - see below

\subsection{Are different traits equivalent?}
%     Can disparity analysis work with any kind of trait?
A common practice in morphological disparity analysis is to recycle data from other studies which can introduce additional problems.
One classic case is using ``cladistic'' data in palaeobiology.
These data are usually collected for phylogenetic analysis and are thus designed to maximise differences among taxa to recover a clear phylogenetic signal.
It can therefore artificially lead to maximised disparity among taxa [Foote - crinoids non-cladistic paper].
A relatively simple way to reduce this specific bias is to select the most pertinent characters for the question from the dataset, rather than using all of the data.
For example, in a study of the disparity of dietary niches, removing non-dietary characters is a good approach.
However, it is important to remember that there is no one-to-one mapping between form and function, and that some characters can have multiple functions (e.g. limb traits can be used for both locomotion and feeding).

\subsection{Are different dataset equivalent?}
%     Do discrete and continuous trait reflect the same thing? If yes, what does that imply?
Morphological disparity analysis can be based on three main types of data:
\begin{enumerate}
    \item discrete data that can either describe a feature (often referred to as ``cladistic''; e.g. the presence or absence of legs, the colour of a tail, etc. [CITE]) or count the number of features (e.g. number of digits, etc. [CITE]);
    \item continuous data that can be either simple measurements (e.g. body length, etc. [CITE]) or more abstracted measures (e.g. geometric morphometrics, mathematical contour description, etc. [CITE]);
    \item any mixture of discrete and continuous data (e.g. ratios [CITE]).
\end{enumerate}
None of these types of data are fundamentally better but each will be more appropriate for specific questions.
For example, when investigating variation within bat wing shapes, both homologous landmarks (for geometric morphometric analysis) or continuous measurements of bones may be appropriate.
On the other hand, if the question focuses on the convergence between bat and bird wings, the homology of the traits to measure is more complex to define and the outline shape of the wings might more readily answer the question.
This again illustrates the importance of framing the question.
In the above example, if the question is whether the aerodynamic properties of wings vary within bats or between bats and birds, the traits collected should reflect these aerodynamic properties (e.g. wingspan, aspect ratio, etc.).
Conversely if the question focuses on phylogenetic convergence aspects between different types of species with active flight, it might be more pertinent to use evolutionary traits that may have facilitated flight within and between both groups (e.g. digit length, integumentary system, etc.).

\subsection{Does the part represent the whole?}
%     Does the part represent the whole?
Melanie's work.

%Bonus
There are many other considerations when choosing the appropriate data to use. 
It is essential to ensure that disparity patterns can be linked to general processes rather than being simply the reflection of idiosyncrasies in the data.
For example, are the observed differences between two groups related to factors unrelated to group designation such as sexual dimorphism?
Are the taxa/groups being compared fairly consistent and truly comparable?
In addition, data can be missing, non-overlapping, hierarchical (e.g. inapplicable), ambiguous, polymorphic or correlated [String of morpho data problems cite].
It is important to ensure that any patterns detected are not merely reflections of these features of the data.

Ultimately, data collection is often constrained by the time available, financial considerations and other unavoidable issues, however, these caveats should be kept in mind during the analyses and subsequent interpretation of the results.
% OU -> And how do these traits need to be analysed?

\section{Disparity is multidimensional}
% IN -> Disparity is multidimensional by essence
% Visualisation problem
\subsection{Visualising the multidimensional problem}
\label{visualisation}
It is important to remember that visual interpretations of multidimensional data can easily be misleading.
For example, it is incorrect to assume that because some groups overlap on a bivariate or three dimensional plot, that they overlap across the whole multidimensional space (especially if the space is not Euclidean).
Conversely, when two groups do not overlap on a two or three dimensional plot, this does not mean they never overlap in the multidimensional space.
Furthermore, multidimensional spaces are not necessary Euclidean.
Such non-Euclidean spaces often have non-intuitive properties, for example straight lines in some dimensions are not straight in the space and, even less intuitively, the distance between points A and B might not be equal to the distance between points B and A [CITE Sylvain].

%     Do we need trait reduction (ordination) or can we do without?

%     What do we measure from these multidimensional spaces?
% OU -> And how do we test these things?

\section{Disparity hypotheses}
% IN -> So how do we go for testing hypotheses in disparity?
%     What are our null expectations?
The final stage of most disparity analyses is statistical testing of the disparity metrics, to determine whether changes thought time or between groups are statistically significant. 
Similar to the points above, the statistical test to apply depends on which question is being tackled with disparity analysis, i.e. which hypothesis is being tested.
It also requires consideration of the data, the ordination method used (or the absence thereof) and the metric used.

%     How do we take phylogeny into account?

Alternatively, ancestral state estimation can be used to provide realistic null expectations from the data.
In disparity-through-time analysis, it is possible to sample many ancestral state estimates on the nodes and along branches to test whether the recovered pattern could be simply explained by the phylogenetic structure itself rather than by morphological changes.
Furthermore, this method can be pertinent for fitting models of evolution to investigate whether the recovered disparity metric can be explained by the tested modes of evolution.
%     How can we analyse it through time?

\section{A disparate future?}
% IN -> The definition of disparity and its uses have broadened since the early disparity start.

As mentioned above, there is no ``one-size-fits-all'' morphological disparity analysis pipeline.
As with any multidimensional analysis, there are many nuances in deciding which data to use and how to analyse it, stemming from the explicit formulation of the disparity hypothesis.
To understand these nuances, it is important to remember that: (1) morphological disparity analyses are not new and much effort has been made in developing methods; and (2) multidimensional analyses are not restricted to morphological data analysis and solutions to specific problems can be found in literature that is not necessary focused on morphology (especially ecology and mathematics).

%     What can we get from other fields to make disparity better
%     But also what can we bring to other fields (discussions with Abigail)

\begin{itemize}
    \item{Which data should we use in disparity analyses?}
    [One line abstract of Melanie's paper on inapplicable data - Needs to be developed].
    Furthermore, some simple methods already exist to account for caveats in data. 
    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
    % NC: This item currently doesn't seem to actually give advice about which data to use...
    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
    [One line abstract of Sylvain's paper on morphospaces plottings].
    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
    One solution to this is to use more simple 2D/3D ordinations.
    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
    \item{Which disparity metrics should we use?}
    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g. in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
    \item{How should we test disparity hypotheses?}
    Finally, it is possible to use non-parametric methods to test specific disparity questions.
    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
\end{itemize}

\section{Introduction}

@@

\begin{itemize}
    \item{Which data should we use in disparity analyses?}
    [One line abstract of Melanie's paper on inapplicable data - Needs to be developed].
    Furthermore, some simple methods already exist to account for caveats in data. 
    For example, to correct for the paucity of the fossil record, one can use methods like rarefaction analysis to take sample size into account [CITE].
    % NC: This item currently doesn't seem to actually give advice about which data to use...
    \item{How should we visualise disparity?} % NC: Tried to make the bullets a bit clearer, might need to edit for correctness and consistency with the rest of the document.
    [One line abstract of Sylvain's paper on morphospaces plottings].
    One major problem of multidimensional analyses is that highly dimensional spaces are hard (or even impossible) to conceptualise.
    One solution to this is to use more simple 2D/3D ordinations.
    For example, using and MDS and reducing that data to only two or three dimensions for visualisation purposes only; or alternatively, reducing the number of variables prior to the analysis specifically for visualisation [e.g Raup-spaces CITE].
    \item{Which disparity metrics should we use?}
    This can be tackled by simply using known metrics with desired properties, proposing new metrics, or (even better) using a combination of metrics that will represent a combination of properties of interest in the morphospace.
    For example, when reducing a multidimensional space to one metric, it would be judicious to indicate what this metric represents in the results (e.g. in a plot or a table, if the metric used was the sum of variances, labelling it as ``size of the group'' rather than ``disparity'').
    This is especially advantageous when using multiple disparity metrics (e.g. ``size of the group'' and ``position of the group'').
    \item{How should we test disparity hypotheses?}
    Finally, it is possible to use non-parametric methods to test specific disparity questions.
    For example, using available software and computer power, it is possible to propose specific null models for the expectation of what disparity would be expected to look like under various null models.
    This has been already successfully used in ecology [CITE Diaz paper] and is currently being implemented in palaeobiology (Mark stuff).
\end{itemize}
